Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation

### 2022-03-25 methods.tex
Heterogenous permittivity - repair equation (8)
Previous wrong commit
added lu formulation

There are several possible boundary integral formulations of Equation \eqref{eq:pbe}.~\cite{SearchCooperWout2022} The simplest form was presented by Yoon and Lenhoff in 1991,~\cite{YoonLenhoff1990} known as the \emph{direct} formulation, which only couples the potential at the interface, and not its derivative.
Applying Green's second identity to Equation \eqref{eq:pbe} we get 
%
\begin{align} \label{eq:volume_potential}
\phi_{1}+ K_{L}^{\Omega_1}(\phi_{1,\Gamma}) -  V_{L}^{\Omega_1} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right) & = \frac{1}{\epsilon_1} \sum_{k=0}^{N_q}  \frac{q_k}{4\pi|\mathbf{r}_{\Omega_1} - \mathbf{r}_k|}  \quad \text{on $\Omega_1$,} \nonumber \\
\phi_{2} - K_{Y}^{\Omega_2}(\phi_{2,\Gamma}) + V_{Y}^{\Omega_2} \left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) & = 0 \quad \text{on $\Omega_2$,}
\end{align}
%
where $\phi_{1,\Gamma} = \phi_1(\mathbf{r}_\Gamma)$ and $\phi_{2,\Gamma} = \phi_2(\mathbf{r}_\Gamma)$ is the potential on $\Gamma$ as we approach from $\Omega_1$ and $\Omega_2$, respectively. $K$ and $V$ are the double- and single-layer potentials for the Laplace (subscript $L$) and Yukawa (subscript $Y$, also known as modified Helmholtz) kernels
%
\begin{align}\label{eq:single_double}
V^\Omega_{L,Y}(\varphi) = \oint_\Gamma g_{L,Y}(\mathbf{r}_\Omega,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'\nonumber\\
K^\Omega_{L,Y}(\varphi) = \oint_\Gamma \frac{\partial g_{L,Y}}{\partial\mathbf{n}'}(\mathbf{r}_\Omega,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}',\nonumber\\
\end{align}
%
where $\varphi(\mathbf{r})$ can be any distribution over $\Gamma$, and 
%
\begin{align}\label{eq:green_func}
g_L(\mathbf{r},\mathbf{r}')=\frac{1}{4\pi|\mathbf{r}-\mathbf{r}'|} \nonumber \\
g_Y(\mathbf{r},\mathbf{r}')=\frac{e^{-\kappa|\mathbf{r}-\mathbf{r}'|}}{4\pi|\mathbf{r}-\mathbf{r}'|}
\end{align}
%
are the free-space Green's function of the Laplace and linearized Poisson-Boltzmann equations, respectively.

In the \emph{direct} formulation, we take the limit of Equation \eqref{eq:volume_potential} as $\mathbf{r}$ approaches $\Gamma$and apply the interface conditions for $\phi$ and $\epsilon\partial\phi/\partial\mathbf{n}$ to get
%
\begin{align} \label{eq:direct}
\frac{\phi_{1,\Gamma}}{2}+ K_{L}^{\Gamma}(\phi_{1,\Gamma}) -  V_{L}^{\Gamma} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right) & = \frac{1}{\epsilon_1} \sum_{k=0}^{N_q}  \frac{q_k}{4\pi|\mathbf{r}_{\Gamma} - \mathbf{r}_k|} \nonumber \\
\frac{\phi_{1,\Gamma}}{2} - K_{Y}^{\Gamma}(\phi_{1,\Gamma}) + \frac{\epsilon_1}{\epsilon_2}V_{Y}^{\Gamma} \left( \frac{\partial}{\partial \mathbf{n}} \phi_{1,\Gamma} \right) & = 0
\end{align}
%
Even though this formulation is poorly conditioned with respect to the mesh size, it can model large systems when it is preconditioned appropriately.\cite{AltmanBardhanWhiteTidor09,WangCooperBetckeBarba2021}

@@

BEM reduces the dimension of the problem by using the boundary integral equation, hence its popularity as a numerical solution of  the Poisson-Boltzmann equation in molecular electrostatics. Unfortunately, such advantage comes with the cost. A fundamental solution must be found before the BEM can be applied. There are many linear problems for which fundamental solutions are not known. For Poisson-Boltzmann equation it is a case if we consider the heterogenous permittivity inside the molecule
   \begin{align} \label{eq:pbe_vp}
\epsilon_1(\mathbf{x}) \nabla^2\phi_1(\mathbf{x}) &= \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2(\mathbf{x})  &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1(\mathbf{x})  = \phi_2(\mathbf{x}),  &\quad \epsilon_1(\mathbf{x})\frac{\partial\phi_1}{\partial\mathbf{n}}(\mathbf{x})  = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}}(\mathbf{x})  \quad \mathbf{x}\in \Gamma. 
\end{align}
For such case finite element methods (FEM) are more suitable.

>>>>

There are several possible boundary integral formulations of Equation \eqref{eq:pbe}.~\cite{SearchCooperWout2022} The simplest form was presented by Yoon ºand Lenhoff in 1991,~\cite{YoonLenhoff1990} known as the \emph{direct} formulation, however, the resulting system is usually ill-conditioned. Here, we use the better conditioned formulation presented by Lu and co-workers.~\cite{LuETal2006,LuETal2009,ZhangETal2019}
Applying Green's second identity to Equation \eqref{eq:pbe} we get 
%
\begin{align} \label{eq:volume_potential}
\phi_{1}+ K_{L}^{\Omega_1}(\phi_{1,\Gamma}) -  V_{L}^{\Omega_1} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right) & = \frac{1}{\epsilon_1} \sum_{k=0}^{N_q}  \frac{q_k}{4\pi|\mathbf{r}_{\Omega_1} - \mathbf{r}_k|}  \quad \text{on $\Omega_1$,} \nonumber \\
\phi_{2} - K_{Y}^{\Omega_2}(\phi_{2,\Gamma}) + V_{Y}^{\Omega_2} \left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) & = 0 \quad \text{on $\Omega_2$,}
\end{align}
%
where $\phi_{1,\Gamma} = \phi_1(\mathbf{r}_\Gamma)$ and $\phi_{2,\Gamma} = \phi_2(\mathbf{r}_\Gamma)$ is the potential on $\Gamma$ as we approach from $\Omega_1$ and $\Omega_2$, respectively. $K$ and $V$ are the double- and single-layer potentials for the Laplace (subscript $L$) and Yukawa (subscript $Y$, also known as modified Helmholtz) kernels
%
\begin{align}\label{eq:single_double}
V^\Omega_{L,Y}(\varphi) = \oint_\Gamma g_{L,Y}(\mathbf{r}_\Omega,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'\nonumber\\
K^\Omega_{L,Y}(\varphi) = \oint_\Gamma \frac{\partial g_{L,Y}}{\partial\mathbf{n}'}(\mathbf{r}_\Omega,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}',\nonumber\\
\end{align}
%
where $\varphi(\mathbf{r})$ can be any distribution over $\Gamma$, and 
%
\begin{align}\label{eq:green_func}
g_L(\mathbf{r},\mathbf{r}')=\frac{1}{4\pi|\mathbf{r}-\mathbf{r}'|} \nonumber \\
g_Y(\mathbf{r},\mathbf{r}')=\frac{e^{-\kappa|\mathbf{r}-\mathbf{r}'|}}{4\pi|\mathbf{r}-\mathbf{r}'|}
\end{align}
%
are the free-space Green's function of the Laplace and linearized Poisson-Boltzmann equations, respectively.

If we take the limit of Eq. \eqref{eq:volume_potential} as $\mathbf{r}$ approaches $\Gamma$, we get
%
\begin{align} \label{eq:direct}
\frac{\phi_{1,\Gamma}}{2}+ K_{L}^{\Gamma}(\phi_{1,\Gamma}) -  V_{L}^{\Gamma} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right) & = \frac{1}{\epsilon_1} \sum_{k=0}^{N_q}  \frac{q_k}{4\pi|\mathbf{r}_{\Gamma} - \mathbf{r}_k|} \nonumber \\
\frac{\phi_{2,\Gamma}}{2} - K_{Y}^{\Gamma}(\phi_{2,\Gamma}) + V_{Y}^{\Gamma} \left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) & = 0
\end{align}
%
The {\it direct} formulation\cite{YoonLenhoff1991} couples the two expressions of Eq. \eqref{eq:direct} on $\Gamma$ with the interface conditions. Even though this formulation is poorly conditioned with respect to the mesh size, it can model large systems when it is preconditioned appropriately.\cite{AltmanBardhanWhiteTidor09,WangCooperBetckeBarba2021} 
The formulation presented by Lu\cite{LuETal2006} takes the normal derivative of \eqref{eq:direct} to couple both $\phi$ and $\partial\phi/\partial\mathbf{n}$ as
%
\begin{align}\label{eq:lu}
    & \frac{\phi_{2,\Gamma}}{2}\left(\frac{\epsilon_1}{\epsilon_2}+1\right) - \left(K_Y^\Gamma - \frac{\epsilon_1}{\epsilon_2}K_L^\Gamma\right)(\phi_{2,\Gamma}) + \left(V_Y^\Gamma - V_L^\Gamma\right)\left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) = \sum_{k=0}^{N_q}  \frac{q_k}{4\pi\epsilon_2|\mathbf{r}_{\Gamma} - \mathbf{r}_k|}
     \nonumber \\
    &\frac{\epsilon_1}{\epsilon_2}\left(W_Y^\Gamma - W_L^\Gamma\right)(\phi_{2,\Gamma}) +  \frac{1}{2}\frac{\phi_{2,\Gamma}}{\partial\mathbf{n}}\left(1+\frac{\epsilon_1}{\epsilon_2}\right) + \left(\frac{\epsilon_1}{\epsilon_2}K_Y^{\prime\Gamma} - K_L^{\prime\Gamma}\right)\left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) = \sum_{k=0}^{N_q}  \frac{\partial}{\partial\mathbf{n}_\mathbf{r}}\left(\frac{q_k}{4\pi\epsilon_2|\mathbf{r}_{\Gamma} - \mathbf{r}_k|}\right)
\end{align}
%
The two new integral operators that appear are defined as 
%
\begin{align}\label{eq:adj_hyp}
K^{\prime\Gamma}_{L,Y}(\varphi) = \oint_\Gamma \frac{g_{L,Y}}{\partial\mathbf{n}}(\mathbf{r}_\Gamma,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'\nonumber\\
W^\Gamma_{L,Y}(\varphi) = - \oint_\Gamma \frac{\partial^2 g_{L,Y}}{\partial\mathbf{n}'\partial\mathbf{n}}(\mathbf{r}_\Gamma,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'
\end{align}
%
and are called the {\it  adjoint double layer} and {\it hypersingular} operators, respectively. Note that the potential in Eq. \eqref{eq:lu} is evaluated on the exterior.

Having computed the electrostatic potential with Eq. \eqref{eq:lu}, the reaction potential ($\phi_r$) in Eq. \eqref{eq:energy} by subtracting out the Coulombic component out of $\phi_1$ in Eq. \eqref{eq:volume_potential}. This gives
%
\begin{align} \label{eq:phi_reac}
\phi_{r} = - K_{L}^{\Omega_1}(\phi_{1,\Gamma}) + V_{L}^{\Omega_1} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right)
\end{align}

@@

BEM reduces the dimension of the problem by using the boundary integral equation, hence its popularity as a numerical solution of  the Poisson-Boltzmann equation in molecular electrostatics. Unfortunately, such advantage comes with the cost. A fundamental solution must be found before the BEM can be applied. There are many linear problems for which fundamental solutions are not known. For Poisson-Boltzmann equation it is a case if we consider the heterogenous permittivity inside the molecule
   \begin{align} \label{eq:pbe_vp}
\nabla \cdot \left(\epsilon_1(\mathbf{x}) \nabla \phi_1(\mathbf{x})\right) &= \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2(\mathbf{x})  &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1(\mathbf{x})  = \phi_2(\mathbf{x}),  &\quad \epsilon_1(\mathbf{x})\frac{\partial\phi_1}{\partial\mathbf{n}}(\mathbf{x})  = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}}(\mathbf{x})  \quad \mathbf{x}\in \Gamma. 
\end{align}
For such case finite element methods (FEM) are more suitable.
